---
title: "bioconductor_singlecell"
author: "Ahmed"
date: "09/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(DropletUtils)
```
#Load the data

```{r}
pbmc <- read10xCounts(samples = c('pbmc1k'="./data/filtered_feature_bc_matrix/"))
```

```{r}
colData(pbmc)
```
Compute and view quality control

```{r}
library(scater)

qc_feature <- perFeatureQCMetrics(pbmc)
qc_feature
```

```{r}
qc_cell <- perCellQCMetrics(pbmc)
head(qc_cell$sum, n=10)
qc_cell

plot1 <- ggplot(as.data.frame(qc_cell), aes(percent_top_50))+
    geom_histogram()
    
plot1

plot2 <- ggplot(as.data.frame(qc_cell),aes(sum,percent_top_50))+
    geom_point()+
    geom_vline(xintercept=2000)+
    geom_hline(yintercept = 60)

plot2

plot3 <- ggplot(as.data.frame(qc_cell),aes(sum,detected,color=percent_top_50))+
    geom_point()
plot3

#plot1+plot2+plot3
```
#Identify mt genes

```{r}
mt_gene <- grep('^MT-',rowData(pbmc)[,'Symbol'])
```

# Use addPerCellQC() to add the mitochondrial genes in the pbmc dataset
```{r}
pbmc <- addPerCellQC(pbmc,subsets=list(MT=mt_gene))

colData(pbmc)

plot4 <- ggplot(as.data.frame(colData(pbmc)),aes(sum,detected,color=subsets_MT_percent))+
    geom_point()
plot4
```


# QC for each feature
```{r}
qc_feature <- perFeatureQCMetrics(pbmc)
qc_feature
```
#To subset:
```{r}
metadata(pbmc)

keep_genes <- 1:5
keep_cells <- 10:50

pbmc_subset <- pbmc[keep_genes, keep_cells]

rowData(pbmc_subset)
colData(pbmc_subset)

rownames(pbmc) <- rowData(pbmc)$Symbol
```
#Normalisation:

```{r}

library(scran)
pbmc <- logNormCounts(pbmc)

dec.pbmc <- modelGeneVar(pbmc)

fit.pbmc <- metadata(dec.pbmc)

mean(pbmc)

plot(fit.pbmc$mean, fit.pbmc$var, xlab="Mean of log-expression", ylab="variance of log-expression")
curve(fit.pbmc$trend(x), col="dodgerblue", add=TRUE, lwd=2)
```
#A more complicated way of doing the step above
```{r}
#df <- data.frame(
    #var = rowVars(as.matrix(assay(pbmc, "logcounts"))),
    #mean = rowMeans(as.matrix(assay(pbmc, "logcounts")))
#)

#ggplot(df, aes(mean, var))+
    #geom_point()+
    #geom_smooth(color="blue", method= "loess", formula = y ~ x)+
    #theme_bw()
```
#Feature selection: select featurs for downstream analysis i.e. highly variable genes

```{r}
dec.pbmc

hist(dec.pbmc$total - (dec.pbmc$tech + dec.pbmc$bio), breaks = 1000)

ggplot()+
    geom_point(aes(mean, bio), as.data.frame(dec.pbmc), color="blue")+
    geom_hline(yintercept = 0, linetype = "dashed", size = 2, color = "green")
```
#Identify highly variable genes
#biological variance can be due to the different types of cells
#variation threshld can be changed accordingly i.e. if you didn't get what to expect in clustering

```{r}
hvgs <- getTopHVGs(dec.pbmc, var.threshold = 0.05, fdr.threshold = 0.05)
#View(hvgs)
View(hvgs)

dec.pbmc$gene_id <- rownames(dec.pbmc)
View(as.data.frame(dec.pbmc))
#View(as.data.frame(dec.pbmc))

ggplot()+
    geom_point(aes(mean, bio, color=gene_id %in% hvgs), as.data.frame(dec.pbmc))+
    geom_hline(yintercept = 0, linetype = "dashed", size = 1, color = "green")
```
#Dimensionality reduction
#Run PCA for the selected genes only

```{r}
pbmc <- runPCA(pbmc, subset_row= hvgs)
pbmc
```
plot the PCA using elbow plot
```{r}
str(reducedDim(pbmc, "PCA"))


percent.var <- attr(reducedDim(pbmc), "percentVar")
plot(percent.var, log="y", xlab="PC", ylab="var")
```

```{r}
pbmc <- runUMAP(pbmc, dimred = "PCA", n_dimred = 20)
reducedDim(pbmc, "UMAP")

pbmc <- runUMAP(pbmc, dimred = "PCA", n_dimred = 20)


plotReducedDim(pbmc, dimred = "PCA", colour_by = "sizeFactor")

plotReducedDim(pbmc, dimred = "UMAP", colour_by = "sizeFactor")
```
#Clustering

```{r}
g <- buildSNNGraph(pbmc, use.dimred = 'PCA')
g
clust <- igraph::cluster_walktrap(g)$membership
table(clust)
```

```{r}

pbmc <- runTSNE(pbmc, dimred = "PCA", n_dimred = 20)

pbmc
colLabels(pbmc) <- factor(clust)
plotReducedDim(pbmc, "TSNE", colour_by="label")

```
#Clustering cells after using a variable number of PCs

```{r}
output <- scran::getClusteredPCs(reducedDim(pbmc, "PCA"))
output
```
#see what it is suggesting:

```{r}
metadata(output)$chosen
```
the answer is 20 PCs

```{r}
ggplot(as.data.frame(output), aes(n.pcs, n.clusters))+
    geom_point()+
    geom_line()
```
the graph above explains why 20 was the chosen answer. We detect more clusters with 20 PCs
Next, we need to run the PCA with the ideal number of clusters

```{r}
pbmc <- runPCA(pbmc, subset_row=hvgs, ncomponents=20)
dim(reducedDim(pbmc, "PCA"))
```





```{r}
g <- buildSNNGraph(pbmc, use.dimred = "PCA")
g

cluster_walktrap <- igraph::cluster_walktrap(g)
table(cluster_walktrap$membership)


library(igraph)



```

```{r}
plotUMAP(pbmc, colour_by="label")
```

```{r}
out <- scran::findMarkers(pbmc, groups=pbmc$label, test.type="wilcox")
out
```

