---
title: "tidyverse exercise"
author: "Ahmed"
date: "25/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=FALSE}
#this code chunck will not be included in the report

```
Load tidyverse

```{r}
library(tidyverse)
```
x dplyr::filter() masks stats::filter()
x dplyr::lag()    masks stats::lag()
This means the dplyr functions will override stats ones. To use stats one it has to be specified (stats::..etc)

```{r}
getwd()
```

```{r}
counts_table <- read_tsv("data/obds_countstable.tsv.gz")

```
```{r}
sample_table <- read_tsv("data/obds_sampletable.tsv")
```
Tidy count
Three columns - Geneid, sample, count

```{r}
counts_table_tidy <- counts_table %>%
  pivot_longer(-Geneid, names_to = "samples", values_to = "count")
```

Join with gene info to get mgi_symbol

```{r}
library(biomaRt)
#download the ensembl database
ensembl <- useMart("ensembl")
#download a list of datasets 
listDatasets(ensembl)

ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")

#see which attributes are available
listAttributes(ensembl)

#now we want to match the ensemble_gene_id (from our table) to mgi_symbol from the database
#we will have a table of gene names
gene_names <- getBM(c("ensembl_gene_id", "mgi_symbol"), filters = "ensembl_gene_id", values= counts_table$Geneid, mart= ensembl)
```
There is another package to check if we have 1:1 mapping of geneID to name

```{r}
length(unique(gene_names$ensembl_gene_id))
nrow(gene_names)
```
join the counts_table_tidy with gene_names

```{r}
counts_table_tidy <- counts_table_tidy %>%
  left_join(gene_names, by = c("Geneid" = "ensembl_gene_id"))
```

Are there any unnamed genes?

```{r}
sum(is.na(counts_table_tidy))/12
```
this means that there are 163 genes who have ensembl ID but no a geneID

Q - Tidy metadata:
-One variable per column
-Don't need species and library_layout columns

```{r}
metadata_table <- sample_table %>%
  separate(sample_title, c("genotype", "to_merge", "cell_type", "replicate_number"), sep = "_")
```
we had to separate to 4 columns but we need 3 (3 undescore separators)
We need to merge 2 gene name columns
```{r}
metadata_table <- metadata_table %>%
  unite(col = genotype, genotype, to_merge, sep = "_")
```
remove species and library_layout columns

```{r}
metadata_table <- metadata_table %>%
  dplyr::select(-species, -library_layout)
```

Merge everything in one chunk:

```{r}
# metadata_table <- sample_table %>%
#   separate(sample_title, c("genotype", "to_merge", "cell_type", "replicate_number"), sep = "_") %>%
#   unite(col = genotype, genotype, to_merge, sep = "_") %>%
#   dplyr::select(-species, -library_layout)
```

Add metadata table to table with counts and gene info

```{r}
counts_table_tidy <- counts_table_tidy %>%
  left_join(metadata_table, by= c("samples" = "Sample_accession"))
```

Calculate counts per million (CPM) - use group_by() and mutate()

```{r}
counts_table_tidy <- counts_table_tidy %>%
  mutate(CPM = count*1000000/read_count)
```
Also calculate log2(CPM + 1)

```{r}
counts_table_tidy <- counts_table_tidy %>%
  mutate(log_transformed = log2(CPM+1))
```

Use group_by() and summarise()

```{r}
#we are removing the columns for practice only, to practive using group_by and summarise
counts_table_tidy <- counts_table_tidy %>%
  dplyr::select(Geneid:replicate_number)
```


```{r}
#create total counts per sample, as a loop!
counts_table_tidy <- counts_table_tidy %>%
  group_by(samples) %>%
  mutate(total_counts = sum(count))
```
checking if we have 12 unique values

```{r}
length(unique(temp$total_counts))
```
```{r}
counts_table_tidy <- counts_table_tidy %>%
  mutate(CPM=count*1000000/total_counts)
```
#the above gave us CPM
Now the log2

```{r}
counts_table_tidy <- counts_table_tidy %>%
  mutate(log_transformed = log2(CPM+1))
```

Plot with ggplot using geom_col()
#In order to do that, we need to summarise the data frst to have sample name on x axis and counts on y axis



```{r}
#test goup_by function
plotting_table <- counts_table_tidy %>%
  group_by(samples)
```

```{r}
plotting_table <- counts_table_tidy %>%
  group_by(samples) %>%
  summarise(total_counts = sum(count))
```
Now to plotting

```{r}
plot1 <- ggplot(plotting_table, aes(x=samples, y=total_counts, col=samples))+
  geom_col()+
  theme(axis.text.x = element_text(size=10, angle= 90, hjust=1, color = "red"))
plot1
```
#group by gene
#then count how many genes have 0 count
```{r}
temp <- counts_table_tidy %>%
  group_by(Geneid) %>%
  summarise(total_counts_per_gene = sum(count)) %>%
  filter(total_counts_per_gene == 0) %>%
  tally()
```
#Draw a density plot of log2(CPM + 1) for all genes

```{r}
plot2 <- ggplot(counts_table_tidy, aes(x=log_transformed, col=samples))+
  geom_density()+
  theme(axis.text.x = element_text(size=10, angle= 90, hjust=1, color = "red"))

plot2
```
#Filter out genes that have low expression in 3 or fewer samples
-For low expression use CPM < 0.5

#filter() to include he row we want. Pull is to generate a vector of the lsit of genes we want to include
```{r}
genes_to_keep <- counts_table_tidy %>%
  group_by(Geneid) %>%
  filter(CPM >= 0.5) %>%
  tally() %>%
  filter(n > 3) %>%
  pull(Geneid)

length(genes_to_keep)
```
```{r}
filtered_counts_table <- counts_table_tidy %>%
  filter(Geneid %in% genes_to_keep)
nrow(filtered_counts_table)
nrow(counts_table_tidy)
```
#another way

```{r}
filtered_counts_table <- counts_table_tidy %>%
  group_by(Geneid) %>%
  filter(sum(CPM>=0.5)>3)

nrow(filtered_counts_table)
```
#density plot on the filtered table

```{r}
plot3 <- ggplot(filtered_counts_table, aes(x=log_transformed, col=samples))+
  geom_density()+
  theme(axis.text.x = element_text(size=10, angle= 90, hjust=1, color = "red"))
plot3
```

```{r}
plot3
```


