---
title: "Integration"
author: "Ahmed"
date: "11/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import 2 RDS files

```{r}
library(Seurat)

pbmcv2 <- readRDS(file="data/pbmcv2_filtered.rds")
pbmcv3 <- readRDS(file = "data/pbmcv3_filtered.rds")

View(v2@meta.data)
View(v3@meta.data)
```

Create a merged Seurat object

```{r}
pbmc.combined <- merge(pbmcv2, y=pbmcv3, add.cell.ids = c("v2", "v3"), project = "PBMC")

pbmc.combined
```
#Normalisation

```{r}
pbmc.combined <- NormalizeData(object = pbmc.combined, normalization.method = "LogNormalize", scale.factor = 10000)
```

```{r}
pbmc.combined[["percent.mt"]] <- PercentageFeatureSet(pbmc.combined, pattern = "^MT-")

```
#Visualise QC metrics as violin plot

```{r}
VlnPlot(pbmc.combined, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
#Find variable genes

```{r}
pbmc.combined <- FindVariableFeatures(pbmc.combined, selection.method = "vst", nfeatures = 2000)

#top 10 highly variable genes:

top10 <- head(VariableFeatures(pbmc.combined), 10)
top10

#plot variables genes with labels

plot1 <- VariableFeaturePlot(pbmc.combined)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1
```
#Scale the data. get all genes names first, then scale the data so that mean expression is 0 and variance is 1.

```{r}
all.genes <- rownames(pbmc.combined)
pbmc.combined <- ScaleData(pbmc.combined, features = all.genes)


```
Perform linear dimensional reduction

```{r}
pbmc.combined <- RunPCA(pbmc.combined, features = VariableFeatures(object = pbmc.combined))
```
```{r}

```

pbmc.combined <- RunUMAP(pbmc.combined, features = VariableFeatures(object = pbmc.combined))

#Visualise the results

```{r}
DimPlot(pbmc.combined)

```



```{r}
DimHeatmap(pbmc.combined, dims = 1, cells = 500, balanced = TRUE)
```

```{r}
VizDimLoadings(pbmc.combined, dims=1:2, reduction = "pca")
```

```{r}
ElbowPlot(pbmc.combined)
```
#Clustering

```{r}
pbmc.combined <- FindNeighbors(pbmc.combined, dims = 1:10)
pbmc.combined <- FindClusters(pbmc.combined, resolution = 0.5)
```
#Determine the dimensionality of the dataset

```{r}
pbmc.combined <- JackStraw(pbmc.combined, num.replicate = 100)
pbmc.combined <- ScoreJackStraw(pbmc.combined, dims = 1:20)
```
Plot JackStraw

```{r}
JackStrawPlot(pbmc.combined)
```

```{r}
pbmc.combined <- RunUMAP(pbmc.combined, features = VariableFeatures(object = pbmc.combined))

DimPlot(pbmc.combined, reduction = "umap", group.by= "orig.ident", label = TRUE)

clustermarkers1 <- FindMarkers(pbmc.combined, ident.1 = 2, min.pct = 0.25)

head(clustermarkers1)
```
Exercise 2:
#Integration method 1:

```{r}

library(Seurat)
library(SeuratData)
library(patchwork)


pbmc.list <- SplitObject(pbmc.combined, split.by = "orig.ident")
names(pbmc.list)


features <- SelectIntegrationFeatures(object.list = pbmc.list)

anchors <- FindIntegrationAnchors(object.list = pbmc.list, anchor.features = features)
anchors

integrated_pbmc <- IntegrateData(anchorset = anchors)

DefaultAssay(integrated_pbmc) <- "integrated"
```
#Run the workflow
#Scaling

```{r}

integrated_pbmc <- ScaleData(integrated_pbmc, npcs=30, verbose = FALSE)

#run PCA
integrated_pbmc <- RunPCA(integrated_pbmc, features = VariableFeatures(object = integrated_pbmc))

```
plot

```{r}
DimPlot(integrated_pbmc)
```


```{r}
integrated_pbmc <- RunUMAP(integrated_pbmc, features = VariableFeatures(object = integrated_pbmc))

DimPlot(integrated_pbmc, reduction = "umap", group.by= "orig.ident", label = TRUE)
```
#Perform clustering

```{r}
integrated_pbmc <- FindNeighbors(integrated_pbmc, dims = 1:10)
integrated_pbmc <- FindClusters(integrated_pbmc, resolution = 0.5)
```

```{r}
DefaultAssay(integrated_pbmc) <- "RNA"

clustermarkers2 <- FindMarkers(integrated_pbmc, ident.1 = 2, min.pct = 0.25)

head(clustermarkers2, n=10)
```
BiocManager::install('multtest')

```{r}

BiocManager::install('multtest')
install.packages('metap')
nk.markers <- FindConservedMarkers(integrated_pbmc, ident.1 = 6, grouping.var = "orig.ident", verbose = FALSE)
head(nk.markers)

FeaturePlot(integrated_pbmc, features = c("IGHG1", "TNFRSF13B", "CLECL1", "IGHG3"))
```

