---
title: "DEseq2 exercise"
author: "Ahmed"
date: "03/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
renv::restore()
```


open required packages
```{r}
library(DESeq2)
library(tidyverse)
```
Read in tables

```{r}
counts_table <- read.table("data/obds_countstable.tsv.gz", header=TRUE, row.names = 1)

sample_table <- read.table("data/obds_sampletable.tsv", header=TRUE, row.names = 1, sep="\t")

#DESeqDataSetFromMatrix()
```
#DeSeqDataSetFromMatrix() needs a matrix 

```{r}
counts_table <- as.matrix(counts_table)
```
The DESeqDataSetFromMatrix() function
we need t remove the number of replicates because we will be comparing KO in CD4 and CD8

```{r}
sample_table <- separate(sample_table, sample_title, into=c("to_merge1", "to_merge2", "to_merge3", "replicate"), sep = "_")
sample_table <- unite(sample_table, "egr_ko_cell", to_merge1:to_merge3, sep="_")
```

Set the reference

```{r}
sample_table$egr_ko_cell <- factor(sample_table$egr_ko_cell, levels = c("Egr2/3_DKO_CD8", "Egr2/3_DKO_CD4", "Egr2_Kin_CD8", "Egr2_Kin_CD4"))
```
Put the data into a DESeq2 data object
use colData to se the column info
```{r}
dds <- DESeqDataSetFromMatrix(counts_table, sample_table, ~egr_ko_cell)
colData(dds)
levels(dds$egr_ko_cell)
```
#4.Access the design formula, counts matrix and sample information from dds


```{r}
assays(dds)$counts
#assay(dds, "counts") -> another way
#show the design formula
design(dds)
colData(dds)
```
```{r}
dds <- estimateSizeFactors(dds)
dds
sizeFactors(dds) #balances out counts based on library size e.g. first sample counts will be timed by 1.0350641
class(sizeFactors(dds))
```
#Generate a bar plot of the size factors for each sample, coloured by condition/group
make anew table (df) with all these data

```{r}
plotting_table <- data.frame(sample = row.names(colData(dds)), size_factors= sizeFactors(dds), condition=colData(dds)$egr_ko_cell)
```


```{r}
library(ggplot2)
ggplot(plotting_table, aes(x=sample, y=size_factors, fill=condition))+
    geom_col()+
    theme(axis.text.x = element_text(angle = 60, hjust=1))
```
Obtain dispersion estimates for each gene – estimateDispersions()

```{r}
dds <- estimateDispersions(dds)
head(assays(dds)$mu)
dispersions(dds)
```

Plot the per-gene dispersion estimates (DESeq2 has a helper function for this)

```{r}
plotDispEsts(dds)
```
7. Perform the Wald test – nbinomWaldTest()

```{r}
dds <- nbinomWaldTest(dds)
```
#Run the three steps in 1 function DESeq

```{r}
dds <- DESeqDataSetFromMatrix(counts_table, sample_table, ~egr_ko_cell)
dds <- DESeq(dds)
```
9. Access the coefficients of the NB GLM

```{r}
coef_output <- coef(dds)
#to access p values convert it to a data frame
View(as.data.frame(colData(dds)))
p_values <- results(dds, c("egr_ko_cell", "Egr2/3_DKO_CD4", "Egr2/3_DKO_CD8"))
p_values_as_df <- as.data.frame(p_values)
```
10. Plot a histogram of the raw and BH-adjusted p-values – do they lookas expected?

```{r}
p_adj_histogram <- ggplot(p_values_as_df, aes(x=padj))+
    geom_histogram()
p_adj_histogram


```
```{r}
p_histogram <- ggplot(p_values_as_df, aes(x=pvalue))+
    geom_histogram()
p_histogram
```

```{r}
library(cowplot)
plot_grid(p_histogram, p_adj_histogram, labels = c("A", "B"), ncol = 2, nrow = 1)
```
12. Generate an MA plot of the log2 FC values for all genes

```{r}
plotMA(p_values)
```
13. Shrink the log2 FC values using the normal, apeglm and ashr methods


```{r}
dds_normal <- lfcShrink(dds, contrast = c("egr_ko_cell", "Egr2/3_DKO_CD4", "Egr2/3_DKO_CD8"), type = "normal")
resultsNames(dds)
dds_apeglm <- lfcShrink(dds, coef="egr_ko_cell_Egr2.3_DKO_CD4_vs_Egr2.3_DKO_CD8", type = "apeglm")
dds_ashr <- lfcShrink(dds, contrast = c("egr_ko_cell", "Egr2/3_DKO_CD4", "Egr2/3_DKO_CD8"), type = "ashr")

#14. Generate MA plots of the shrunken log2 FC values – compare methods

plot_grid(plotMA(dds_normal), plotMA(dds_apeglm), plotMA(dds_ashr), ncol = 3) #this generates 3 separate charts

plotMA(dds_normal)
p1 <- recordPlot()    # save current plot in R variable
 
plotMA(dds_apeglm)
p2 <- recordPlot()
 
plotMA(dds_ashr)
p3 <- recordPlot()
 
plot_grid(p1, p2, p3, ncol = 3)

```
15. Generate a results table (one shrinkage method) containing mgi symbols

- Use the org.Mm.eg.db package
- Remove all genes with a padj of NA
- How many Ensembl IDs are not assigned an mgi symbol and how many mgi
symbols are duplicated?
```{r}
library(org.Mm.eg.db) #alternative to BioMart
#use one of the table with the shrunken p values

dds_apeglm_as_df <- as.data.frame(dds_apeglm)

#Remove all genes with NA
dds_apeglm_na <- dds_apeglm_as_df %>%
    filter(!is.na(padj))

#assign geneID to ensemble names. The package is called mapIds
#the first argument is the library that is actually a database
#the second is the ENSEMBL ID

keytypes(org.Mm.eg.db)
temp <- mapIds(org.Mm.eg.db, rownames(dds_apeglm_na), column = "SYMBOL", keytype = "ENSEMBL")
head(temp)
```
```{r}
dds_apeglm_na_geneID <- as.data.frame(temp)

#make a new column in the dataframe

dds_apeglm_na_geneID$ENSEMBL_ID <- rownames(dds_apeglm_na_geneID)

dds_apeglm_na$ENSEMBL_ID <- rownames(dds_apeglm_na_geneID)

dds_apeglm_joined <- left_join(dds_apeglm_na_geneID, dds_apeglm_na, by = "ENSEMBL_ID")

merged_table <- dds_apeglm_joined[,-c(3,5:6)]
write.csv(merged_table, file = "KO_CD8_vs_KO_CD4.csv")
#this will be the background set in gene enrichment analysis

dds_apeglm_filtered <- dds_apeglm_joined %>%
    filter(padj<0.05 & log2FoldChange>1)

write.csv(dds_apeglm_filtered, file = "sig_DE_genes.csv")
 
```

