---
title: "Chick weight exerceise"
author: "Ahmed"
date: "02/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
View(ChickWeight)
```
Fit a linear model to measure the effect of Time and Diet in the ChickWeight data set

```{r}
lm(weight ~ Time + Diet, ChickWeight)
```
plot it

```{r}
lm_plot <- ggplot(ChickWeight, aes(Time, weight, color=Diet)) +
    geom_point()+
    stat_smooth(method = "lm")
lm_plot
```

```{r}
lm(weight ~ Time, ChickWeight)
```
#the above eans that the weight at point 0 in time is 27.5 (grams) and for every increase in time unit (day) the chicks gained 8.8 grams

```{r}
lm_plot <- ggplot(ChickWeight, aes(Time, weight)) +
    geom_point()+
    stat_smooth(method = "lm")+
    geom_abline(slope = 8.803, intercept = 27.467)
lm_plot
```
# slope is the the rate of change and intercept is the value at point 0
# we can now define the intercept to be diet 1, so all other diets will compare to 1 instead of zero

```{r}
lm(weight ~ Diet, ChickWeight)
```
#The coefficient means that Diet2 is 19.97 grams higher than Doet 1 at any point in time

```{r}
ggplot(ChickWeight, aes(Diet, weight))+
    geom_boxplot()+
    geom_point()
```


Which diet leads to the fastest increase in body weight?
-diet 3
```{r}
lm(weight ~ Diet * Time, ChickWeight)
```


```{r}
lm_plot <- ggplot(ChickWeight, aes(Time, weight, color=Diet)) +
    geom_point()+
    stat_smooth(method = "lm")
lm_plot

```

How much does weight increase per unit of Time for the top diet?
- 4.5811 (Diet1 intercept) + 6.8418 (Diet3 slope)

#for sanity check!:
```{r}
ggplot(ChickWeight, aes(Time, weight, color=Diet)) +
    geom_point()+
    stat_smooth(method = "lm")+
    geom_abline(slope = 6.8418+4.5811, intercept = 30.9310-12.6807)
```

Does the top diet drive an increase in body weight that is significantly faster than the next best diet?
 #to nswer this we need run summary()

```{r}
chickweight_lm_fit <- lm(formula = weight ~ Diet * Time, data = ChickWeight)
summary(chickweight_lm_fit)
```

#One way is to relevel the intercept to diet 4 (2nd highest increase in weight), then re-run the model
```{r}
ChickWeight$Diet <- relevel(ChickWeight$Diet, "4")
chickweight_lm_fit <- lm(formula = weight ~ Time * Diet, data = ChickWeight)
summary(chickweight_lm_fit)

```
#Check if the model is working

```{r}
predict(chickweight_lm_fit, tibble(Time=5, Diet="3"))
```




