---
title: "droplets"
author: "Ahmed"
date: "10/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(DropletUtils)

sce <- read10xCounts("data/pbmc_1k_v3_raw/", sample.names = "PBMC")

View(as.data.frame(colData(sce)))

library(scDblFinder)
```
#Now get rid of empty droplets
Remove cells with no counts

```{r}
assay(sce, "counts")

sums <- colSums(assay(sce, "counts"))

sums

sce <- sce[,sums>0]

sce
```
#emptyDrops needs a numeric object
```{r}
emptyDroplets <- emptyDrops(assay(sce, "counts"))

emptyDroplets

is.cell <- emptyDroplets$FDR <= 0.01
sum(is.cell, na.rm = TRUE)
```
Draw plots comparing empty droplets and other droplets (e.g., library size, mitochondrial content, PCA).
#first we need a dataframe. results are: FLASE is background (empty cell). TRUE is a cell significantly different from background. NA:

```{r}
df <- data.frame(librarysize = colSums(assay(sce, "counts")), empty_droplet = is.cell)
df

ggplot(df, aes(x=empty_droplet, y=librarysize))+
  geom_point()
```
We need to do the same as above for mitochondrial dna. Need to be calculated the same as we've done yesterday
```{r}

```

Import the filtered Cell Ranger matrix for 1,000 cells and compare their set of filtered cells.
#import the filtered matrix

```{r}
sce_filtered <- read10xCounts("data/pbmc_1k_v3_filtered/", sample.names = "PBMC")
sce_filtered
```
```{r}
sce <- sce[,which(is.cell)]
```
compare cells we got with the filtered data

```{r}
barcode_raw <- colData(sce)$Barcode
barcode_filtered <- colData(sce_filtered)$Barcode

length(intersect(barcode_filtered, barcode_raw))
```
Outcome of our filtering method is similar to the filtered data

#Removing doublets

```{r}
sce <- scDblFinder(sce)

sce

View(as.data.frame(colData(sce)))
```


```{r}

df <- data.frame(librarysize = colSums(assay(sce, "counts")), doublets = colData(sce)$scDblFinder.class)
df

ggplot(df, aes(x=doublets, y=librarysize))+
  geom_violin()+
  geom_point()
```
#Remove the doublets from the dataset

```{r}
colData(sce)$scDblFinder.class != "doublet"

sce <- sce[,colData(sce)$scDblFinder.class != "doublet"]

sce
```
#Ambient RNA

```{r}
BiocManager::install("celda")
```
identify contamination with ambient rna

```{r}
sce <- decontX(sce)
```

```{r}
assays(sce)$decontXcounts

View(as.data.frame(colData(sce)))
```


Visualise the UMAP layout created by decontx 

```{r}
reducedDimNames(sce)

class(sce)

plotDimReduceCluster(colData(sce)$decontX_clusters, dim1 = reducedDim(sce)[,"DecontX_UMAP_1"], dim2 = reducedDim(sce)[,"DecontX_UMAP_2"])


plotDimReduceCluster(colData(sce)$decontX_clusters, dim1 = reducedDim(sce)[,"DecontX_UMAP_1"], dim2= reducedDim(sce)[,"DecontX_UMAP_2"])

```
Visualize the percentage of contamination estimated by decontX in each cell. Use the plotDecontXContamination() function

```{r}
plotDecontXContamination(sce)
```
Does any cluster display higher (estimated) levels of ambient RNA?
25%



Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use plotDimReduceFeature()

```{r}
head(rownames(sce))

plotDimReduceFeature(as.matrix(assays(sce)$counts), features = "ENSG00000198851", dim1 = reducedDim(sce)[,"DecontX_UMAP_1"], dim2= reducedDim(sce)[,"DecontX_UMAP_2"])
```




