---
title: "PCA"
author: "Ahmed"
date: "01/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Read the file

```{r}
logcounts <- read.csv("data/logcounts.csv", row.names = 1)
```

```{r}
library(tidyverse)
```
Check the type of the datastructure

```{r}
str(logcounts)
```
Chate the dataframe to a matrix

```{r}
logcounts_matrix <- data.matrix(logcounts, rownames.force = NA)
```

```{r}
str(logcounts_matrix)
```
```{r}
logcounts.pca <- prcomp(logcounts_matrix, center = TRUE, scale. = TRUE)
summary(logcounts.pca)
```

```{r}
str(logcounts.pca)
```
We want to rotate to cluster on the samples not the genes
We need to transpose the data

```{r}
transposed_logcounts <-  t(logcounts_matrix)
str(transposed_logcounts)
```
Then run the PCA again on the transposed data

```{r}
transposed_logcounts.pca <- prcomp(transposed_logcounts, center = TRUE, scale. = FALSE)
str(transposed_logcounts.pca)
```
Now the rotation: has GeneID and PC1, PC2, etc
and the $X given the coordinate of each cell relative to PC1, PC2,..etc

Now we areusing the view funciton to view the coordinates
```{r}
View(transposed_logcounts.pca$x)
```

Draw a scatter plot of the top 2 principal components and colour 

```{r}
library(ggplot2)
cell_metadata <- read.csv("data/cell_metadata.csv")
View(cell_metadata)

```

```{r}
logcounts_pca_as_df <- as.data.frame(transposed_logcounts.pca$x[,1:2])
View(logcounts_pca_as_df)
logcounts_pca_as_df$Sample <- rownames(logcounts_pca_as_df) 
#logcounts_pca_as_df

cellandcounts <- logcounts_pca_as_df %>%
    left_join(cell_metadata, by = c("Sample" = "Sample" ))



infection_plot <- ggplot(data = cellandcounts, aes(x=PC1, y=PC2, col=Infection))+
    geom_point()
status_plot <- ggplot(data = cellandcounts, aes(x=PC1, y=PC2, col=Status))+
    geom_point()
time_plot <- ggplot(data = cellandcounts, aes(x=PC1, y=PC2, col=Time))+
    geom_point()
```
Plotting as a grid

```{r}
library(cowplot)
plot_grid(infection_plot, status_plot, time_plot, labels= c("Infection", "Status", "Time"), ncol = 1, nrow = 3)

```
Draw a density plot

```{r}
infection_dplot <- ggplot(data = cellandcounts, aes(x=PC1, col=Infection))+
    geom_density()
status_dplot <- ggplot(data = cellandcounts, aes(x=PC1, col=Status))+
    geom_density()
time_dplot <- ggplot(data = cellandcounts, aes(x=PC1, col=Time))+
    geom_density()
infection_dplot
status_dplot
time_dplot
```
Facet plots
Draw a density plot, faceted by time and infection, colored by status, to inspect the variance of each
experimental factor on the top principal component

```{r}
facet_plot <- ggplot(cellandcounts, aes(x=PC1, col=Status))+
    geom_density()+
    facet_grid(rows = vars(Infection), cols = vars(Time))
facet_plot

```
Draw a bar plot of variance explained by each principal component. How many principal components
would you keep for downstream analyses?

```{r}
variance_table <- data.frame(variance= (transposed_logcounts.pca$sdev)^2)
variance_table$fraction <- variance_table$variance/sum(variance_table$variance)
PC_numbers <- seq(1,nrow(variance_table))
variance_table$PC <- PC_numbers
variance_table <- variance_table %>%
    select(PC, variance, fraction)
View(variance_table)
```
plotting

```{r}
variance_plotting <- ggplot(variance_table, aes(x=PC, y=fraction))+
    geom_col()+
    xlim(0,15)
variance_plotting
```
Find the names of the top genes associated with the principal components

```{r}
gene_influence <- as.data.frame(transposed_logcounts.pca$rotation)
gene_influence$gene_ID <- row.names(gene_influence)

temp <- gene_influence %>%
    arrange(desc(PC1)) %>%
    select(gene_ID, PC1)
View(temp)
```
Visualise the gene expression value of the gene with the highest loading for PC1 as color on a scatter
plot of PC1 and PC2. What do you observe?

```{r}
ENSG00000172183 <- as.numeric(logcounts["ENSG00000172183",])
ENSG00000172183

cellandcounts$ENSG00000172183 <- ENSG00000172183

gene_expression_plot <- ggplot(cellandcounts, aes(x=PC1, y=PC2, col=ENSG00000172183))+
    geom_point()
gene_expression_plot
```
Compare PCA to t-SNE to UMAP. Note that t-SNE and UMAP should be given the subset of principal
components selected above.

```{r}
library(umap)

subset_logcounts.pca <- transposed_logcounts.pca$x[,1:15]

View(subset_logcounts.pca)
logcount_umap <- umap(subset_logcounts.pca)
logcount_umap_layout <- as.data.frame(logcount_umap$layout)
logcount_umap_layout$Sample <- rownames(logcount_umap_layout)
View(logcount_umap_layout)

logcount_umap_layout_metadata <- logcount_umap_layout %>%
    left_join(cell_metadata)
View(logcount_umap_layout_metadata)

umap_plot <- ggplot(logcount_umap_layout_metadata, aes(x=V1, y=V2, col = Time))+
    geom_point()
umap_plot

```


```{r}
out_kmeans <- kmeans(x= transposed_logcounts.pca$x[,1:25], centers=4)
out_kmeans
```
```{r}
str(out_kmeans)
```
```{r}
ggplot(kmeans_scree, aes(k, withinss_sum)) +
    geom_line(linetype = "F1") +
    geom_point() +
    labs(title = "sum(withinss)") +
    cowplot::theme_cowplot()
```

